---
title: "Quadratic Discriminant Analysis with Heart Disease Dataset"
author: "Data Science Analysis"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
    toc_depth: 3
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
    highlight: tango
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6,
  fig.align = "center"
)

install_if_missing <- function(pkgs) {
  for (p in pkgs) {
    if (!requireNamespace(p, quietly = TRUE)) {
      install.packages(p, repos = "https://cloud.r-project.org", quiet = TRUE)
    }
  }
}

install_if_missing(c("MASS", "ggplot2", "dplyr", "corrplot", "gridExtra", "pROC", "tidyr", "knitr"))

library(MASS)
library(ggplot2)
library(dplyr)
library(corrplot)
library(gridExtra)
library(pROC)
library(tidyr)
library(knitr)

set.seed(16)

if (!dir.exists("images")) dir.create("images")
```

## Introduction

Quadratic Discriminant Analysis (QDA) is a supervised classification technique that models each class with its own covariance matrix, allowing for non-linear decision boundaries. This makes QDA more flexible than LDA when class covariances differ.

- **Goal**: Explore QDA using a simple synthetic example and the heart disease dataset.
- **Outputs**: Figures and metrics summarizing QDA performance.

## Sample Data Demonstration

We first illustrate QDA on two Gaussian classes with different covariance matrices.

```{r sample_data}
# Generate two classes with different covariance matrices
generate_sample_data <- function(n_per_class = 200) {
  mu1 <- c(0, 0)
  mu2 <- c(2, 2)
  Sigma1 <- matrix(c(1.0, 0.6, 0.6, 1.0), 2, 2)
  Sigma2 <- matrix(c(1.8, -0.7, -0.7, 1.0), 2, 2)
  class1 <- MASS::mvrnorm(n = n_per_class, mu = mu1, Sigma = Sigma1)
  class2 <- MASS::mvrnorm(n = n_per_class, mu = mu2, Sigma = Sigma2)
  rbind(
    data.frame(x1 = class1[,1], x2 = class1[,2], class = factor("Class A")),
    data.frame(x1 = class2[,1], x2 = class2[,2], class = factor("Class B"))
  )
}

sample_df <- generate_sample_data()
head(sample_df)
```

```{r sample_qda_plot, fig.width=8, fig.height=6}
qda_fit <- qda(class ~ x1 + x2, data = sample_df)

x1_seq <- seq(min(sample_df$x1) - 1, max(sample_df$x1) + 1, length.out = 200)
x2_seq <- seq(min(sample_df$x2) - 1, max(sample_df$x2) + 1, length.out = 200)
grid <- expand.grid(x1 = x1_seq, x2 = x2_seq)
grid$pred <- predict(qda_fit, grid)$class

p_sample <- ggplot() +
  geom_tile(data = grid, aes(x = x1, y = x2, fill = pred), alpha = 0.2) +
  geom_point(data = sample_df, aes(x = x1, y = x2, color = class), size = 1.8, alpha = 0.8) +
  scale_fill_manual(values = c("Class A" = "#c6dbef", "Class B" = "#fdd0a2"), name = "QDA region") +
  scale_color_manual(values = c("Class A" = "#3182bd", "Class B" = "#e6550d")) +
  labs(title = "Sample Data: QDA Decision Regions",
       x = "x1", y = "x2") +
  theme_minimal() +
  theme(legend.position = "bottom",
        panel.background = element_rect(fill = "white"))

print(p_sample)

ggsave("images/01_sample_data_qda.png", p_sample, width = 8, height = 6, dpi = 300, bg = "white")
```

## Heart Disease Dataset

We now apply QDA to the heart disease dataset.

```{r heart_load}
heart <- read.csv("Heart-Disease-Dataset/heart.csv")
cat("Rows:", nrow(heart), "Cols:", ncol(heart), "\n")
head(heart)
```

### Preprocessing

```{r heart_preprocess}
# Drop known faulty rows if present
if ("ca" %in% names(heart)) heart <- heart[!(heart$ca == 4), ]
if ("thal" %in% names(heart)) heart <- heart[!(heart$thal == 0), ]

# Target to factor
if ("target" %in% names(heart)) {
  heart$target <- factor(heart$target, levels = c(0,1), labels = c("NoDisease", "Disease"))
}

# Integer-coded categoricals (optional)
cat_cols <- intersect(c("sex", "cp", "fbs", "restecg", "exang", "slope", "ca", "thal"), names(heart))
for (cc in cat_cols) heart[[cc]] <- factor(heart[[cc]])

# Quick check
str(heart)
```

### EDA

```{r eda_plots, fig.width=10, fig.height=8}
# Target distribution
p_target <- ggplot(heart, aes(x = target, fill = target)) +
  geom_bar(alpha = 0.9) +
  scale_fill_manual(values = c("NoDisease" = "#6baed6", "Disease" = "#fd8d3c")) +
  labs(title = "Target Distribution", x = "Target", y = "Count") +
  theme_minimal() +
  theme(panel.background = element_rect(fill = "white"), legend.position = "none")

print(p_target)

ggsave("images/02_target_distribution.png", p_target, width = 6, height = 4, dpi = 300, bg = "white")

# Correlation plot of numeric variables
num_cols <- names(heart)[sapply(heart, is.numeric)]
if (length(num_cols) > 1) {
  cor_mat <- suppressWarnings(cor(heart[, num_cols], use = "pairwise.complete.obs"))
  corrplot(cor_mat, method = "color", type = "upper", addCoef.col = "black",
           tl.col = "black", tl.srt = 45,
           title = "Heart Dataset: Numeric Correlation Matrix",
           mar = c(0,0,2,0))
}
```

```{r histograms, fig.width=12, fig.height=8}
sel <- intersect(c("age", "trestbps", "chol", "thalach", "oldpeak"), names(heart))
if (length(sel) > 0) {
  long_df <- tidyr::pivot_longer(heart[, c(sel, "target")], cols = all_of(sel), names_to = "feature", values_to = "value")
  p_hist <- ggplot(long_df, aes(x = value, fill = target)) +
    geom_histogram(bins = 30, alpha = 0.6, position = "identity") +
    facet_wrap(~feature, scales = "free", ncol = 2) +
    scale_fill_manual(values = c("NoDisease" = "#6baed6", "Disease" = "#fd8d3c")) +
    labs(title = "Selected Numeric Features", x = "Value", y = "Frequency") +
    theme_minimal() +
    theme(panel.background = element_rect(fill = "white"))
  print(p_hist)
  ggsave("images/04_numeric_histograms.png", p_hist, width = 10, height = 8, dpi = 300, bg = "white")
}
```

## QDA Modeling

```{r qda_model}
# Predictors: numeric only
numeric_cols <- setdiff(names(heart)[sapply(heart, is.numeric)], c("target"))
stopifnot(length(numeric_cols) >= 2)

set.seed(16)
idx <- sample(seq_len(nrow(heart)), size = floor(0.7 * nrow(heart)))
train <- heart[idx, ]
test  <- heart[-idx, ]

form <- as.formula(paste("target ~", paste(numeric_cols, collapse = " + ")))
qda_model <- qda(form, data = train)

pred_train <- predict(qda_model, train)
pred_test  <- predict(qda_model, test)

acc_train <- mean(pred_train$class == train$target)
acc_test  <- mean(pred_test$class == test$target)

cat(sprintf("Train Accuracy: %.2f%%\n", 100 * acc_train))
cat(sprintf("Test  Accuracy: %.2f%%\n", 100 * acc_test))
```

```{r confusion_matrix, fig.width=8, fig.height=6}
cm <- table(Actual = test$target, Predicted = pred_test$class)
cm_df <- as.data.frame(cm)

p_cm <- ggplot(cm_df, aes(x = Predicted, y = Actual, fill = Freq)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "red", name = "Count") +
  geom_text(aes(label = Freq), color = "black", size = 4) +
  labs(title = sprintf("QDA Confusion Matrix (Test) — Acc: %.1f%%", 100 * acc_test),
       x = "Predicted", y = "Actual") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.background = element_rect(fill = "white"))

print(p_cm)

ggsave("images/05_confusion_matrix.png", p_cm, width = 7, height = 5.5, dpi = 300, bg = "white")
```

```{r roc_curve, fig.width=7, fig.height=6}
if (!is.null(pred_test$posterior)) {
  pos_prob <- pred_test$posterior[, "Disease"]
  y_true <- as.numeric(test$target == "Disease")
  roc_obj <- pROC::roc(response = y_true, predictor = pos_prob, quiet = TRUE)
  auc_val <- as.numeric(pROC::auc(roc_obj))

  roc_df <- data.frame(tpr = roc_obj$sensitivities, fpr = 1 - roc_obj$specificities)
  p_roc <- ggplot(roc_df, aes(x = fpr, y = tpr)) +
    geom_line(color = "#2c7fb8", size = 1) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray60") +
    coord_equal() +
    labs(title = sprintf("ROC Curve (AUC = %.3f)", auc_val), x = "False Positive Rate", y = "True Positive Rate") +
    theme_minimal() +
    theme(panel.background = element_rect(fill = "white"))

  print(p_roc)
  ggsave("images/06_roc_curve.png", p_roc, width = 6, height = 5, dpi = 300, bg = "white")
}
```

## Discussion

- QDA models class-specific covariance, enabling curved boundaries and improved performance when class covariances differ.
- For the heart dataset, key numeric predictors (e.g., age, thalach, oldpeak) contribute to separability, but performance depends on train/test split.

## References

- MASS package documentation for QDA implementation
- Heart disease dataset (Cleveland) data dictionary from Kaggle

